Relative parabolic factorizations of Chevalley groups

### 2015-04-30 bass-kolster.tex
Rewritten BK and stability conditions

\begin{thm}\label{thm:BassKolster}
Let $\Phi$ be a classical root system of rank $\geqslant2$, let $R$ be arbitary commutative ring and $I$ be an ideal, satisfying the following condition:
\begin{itemize}
\item $\Phi=\rA_\ell$: $\sr(I)\leqslant\ell$,
\item $\Phi=\rB_\ell$: $\asr(R,I)\leqslant\ell-1$,
\item $\Phi=\rC_\ell$: $\sr(I)\leqslant2\ell$,
\item $\Phi=\rD_\ell$: $\asr(R,I)\leqslant\ell-1$.
\end{itemize}
Then the principal congruence subgroup $\G(\Phi,R,I)$ admits the following relative version of Bass---Kolster decomposition:
\[\G(\Phi,R,I)=\U(\Sigma^+,I)\ \U(\Sigma^-,I)\cdot \Z\cdot \U(\Sigma^-,I)\ \U(\Sigma^+,I)\cdot \G(\Delta,R,I), \]
where $\Sigma^\pm=\Phi^\pm\setminus\Delta$ and $\Z=\{ z_{-\alpha_\mathrm{max}}(r,1)\mid r\in I \}$.
\end{thm}
\begin{proof}
Let $g$ be an element of $\G(\Phi, R, I)$ and $e_1$ be the highest weight vector of the standard vector representation of $\G(\Phi, R, I)$.
It is clear that vector $v=g\cdot e_1$ is $I$-unimodular.
%TODO: Mention Chevalley---Matsumoto lemma in the introduction
In view of Chevalley---Matsumoto lemma it suffices to show that $\Z \cdot \U(\Phi^-, I) \cdot \U(\Phi^+, I) \cdot g$ contains an element $g$ whose matrix entry at position $(1,1)$ is equal to $1$.
We will need to analyze several cases.

\textsc{Case $\Phi=\rA_\ell$:}
Denote $v=(1+r_1,r_2,\ldots,r_\ell,r_{\ell+1})^t\in\Ums(\ell+1,R,I)$.
One can add the last entry to the first $\ell$ with some coefficients from $I$ to make then form a unimodular column of length $\ell$.
Then $\ell$ lower transformations (with coefficients from $I$) can make the last entry equal $r_1$.
Now apply $z_\gamma(-r_1,1)$, $\gamma=-\alpha_{\max}$:
\begin{multline*}
(1+r_1,r_2,\ldots,r_\ell,r_1)^t\xmapsto{x_{-\gamma}(-1)}(1,r_2,\ldots,r_\ell,r_1)^t\xmapsto{x_\gamma(-r_1)}\\\longmapsto(1,r_2,\ldots,r_\ell,0)^t\xmapsto{x_{-\gamma}(1)}(1,r_2,\ldots,r_\ell,0)^t.
\end{multline*}
Now $\ell-1$ lower transformations eliminates all the remaining entries in the first column, while $\ell$ more upper ones will make the first row trivial.

\textsc{Case $\Phi=\rC_\ell$:}
Denote $v=(v^+,v^-)^t\in\Ums(2\ell,R,I)$. Find $c_1,\ldots,c_{-2}\in I$ such that $(v_1+c_1v_{-1},\ldots,v_{-2}+c_{-2}v_{-1})=(v_1',\ldots,v_{-2}')$ is unimodular.

@@

Now apply $2\ell-2$ lower transformation to obtain zeroes on positions $i=2,\ldots,-2$, that will also add $\sum_{i=2}^\ell (v_i+v_{-i})$ to $v_{-1}$,
so by one more transformation one can make the first column trivial. 
The same can be done with the first row, so in total we used $8\ell-3$ transformations.

@@

>>>>

\begin{thm}\label{thm:BassKolster}
Let $\Phi$ be a classical root system of rank $\ell\geqslant2$, let $R$ be arbitary commutative ring and $I$ be an ideal, satisfying one of the following assumptions:
\newline \indent \begin{tabular}{l l l}      
\textbullet & $\Phi = \rA_\ell$, $\ell\geq 2$, & $\sr(I) \leqslant \ell$; \\     
\textbullet & $\Phi = \rB_\ell$, $\rD_\ell$ $\ell\geq 3$,  & $\asr(I) \leqslant \ell-1$; \\ 
\textbullet & $\Phi = \rC_\ell$, $\ell\geq 2$, & $\sr(I) \leqslant 2\ell-1$; \\ 
\end{tabular}

\noindent Then the principal congruence subgroup $\G(\Phi,R,I)$ admits the following relative version of Bass---Kolster decomposition:
\[\G(\Phi,R,I)=\U(\Sigma_1^-\setminus\{-\alpha_\mathrm{max}\},I)\cdot \U(\Sigma_1^+,I)\cdot \Z\cdot \U(\Sigma_1^-,I)\cdot \U(\Sigma_1^+,I)\cdot \G(\Delta_1,R,I), \]
where $\Z=\left\{ z_{-\alpha_\mathrm{max}}(r,1)\mid r\in I \right\}$.
\end{thm}
\begin{proof}
Let $g$ be an element of $\G(\Phi, R, I)$. It is clear that vector $v=g\cdot v^+$ is $I$-unimodular.
In view of relative Chevalley---Matsumoto lemma it suffices to show that $\Z \cdot \U(\Phi^-, I) \cdot \U(\Phi^+, I) \cdot g$ contains an element $g$ whose matrix entry at position $(1,1)$ is equal to $1$.
We will need to analyze several cases.

\textsc{Case $\Phi=\rA_\ell$:}
Denote $v=(1+r_1,r_2,\ldots,r_\ell,r_{\ell+1})^t\in\Ums(\ell+1,R,I)$.
One can add the last entry to the first $\ell$ with some coefficients from $I$ to make them form a unimodular column of length $\ell$.
Then $\ell$ lower transformations (with coefficients from $I$) can make the last entry equal $r_1$.
Now apply $z_\gamma(-r_1,1)$, $\gamma=-\alpha_{\max}$:
\begin{multline*}
(1+r_1,r_2,\ldots,r_\ell,r_1)^t\xmapsto{x_{-\gamma}(-1)}(1,r_2,\ldots,r_\ell,r_1)^t\xmapsto{x_\gamma(-r_1)}\\\longmapsto(1,r_2,\ldots,r_\ell,0)^t\xmapsto{x_{-\gamma}(1)}(1,r_2,\ldots,r_\ell,0)^t.
\end{multline*}

\textsc{Case $\Phi=\rC_\ell$:}
Denote $v=(v^+,v^-)^t\in\Ums(2\ell,R,I)$. 
Find $c_1,\ldots, c_\ell, c_{-\ell},\ldots, c_{-2}\in I$ such that $(v_1+c_1v_{-1},\ldots,v_{-2}+c_{-2}v_{-1})=(v_1',\ldots,v_{-2}')$ is $I$-unimodular.

@@

@@

\begin{rem} %TODO:
\textbf{TODO: Calculate the exact number of $z_\alpha$'s involved in the decomposition}

\textsc{$\Phi=\rA_\ell$}
Now $\ell-1$ lower transformations eliminates all the remaining entries in the first column, while $\ell$ more upper ones will make the first row trivial.

\textsc{$\Phi=\rC_\ell$}
Now apply $2\ell-2$ lower transformation to obtain zeroes on positions $i=2,\ldots,-2$, that will also add $\sum_{i=2}^\ell (v_i+v_{-i})$ to $v_{-1}$, 
so by one more transformation one can make the first column trivial. 
The same can be done with the first row, so in total we used $8\ell-3$ transformations.

\end{rem}

### 2015-04-30 dennis-vaserstein.tex
Rewritten BK and stability conditions

\begin{thm}\label{theorem:relative_dv}
The relative elementary subgroup $\E(\Phi, R, I)$ coincides with $A_{rs}$ under the following assumptions on $(R, I)$.
  \begin{center}
    \begin{tabular}{| l | l | l | l | l |} \hline
    № & $\Phi$ & $(s,r)$ & ring condition \\ \hline
    1. & $\rA_\ell$, $\ell\geq 2$ & $(1, \ell)$ & $\sr(I) \leq \ell-1$ \\ \hline
    2. & $\rB_\ell$, $\ell\geq 3$ & $(1, \ell)$ & $\sr(I) \leq \ell-1$ \\ \hline
    3. & $\rD_\ell$, $\ell\geq 4$ & $(1, \ell)$ & $\sr(I) \leq \ell-2$ \\ \hline    
    4. & $\rE_\ell$, $\ell=6,7$ & $(\ell, 2)$ & $\sr(I) \leq \ell-3$ \\ \hline     
    5. & $\rE_\ell$, $\ell=6,7$ & $(\ell, 1)$ & $\asr(R, I)\leq \ell-2$ \\ \hline    
    \end{tabular} \end{center} 
\end{thm}
The proof of Theorem~\ref{theorem:relative_dv} occupies the rest of this section.

@@

\begin{proof}[Proof of Theorem~\ref{theorem:relative_dv}]
We first show that under specified assumptions on $(R, I)$ one can meet the conditions of Lemma~\ref{lemma:Stein_reduction}.
Consider the following two subsets of $\Lambda(\pi)$:
$$\Gamma = \varpi_s- (\Sigma_s^+\cap \Delta_r),\quad \Gamma_0 = \{\lambda \in \Gamma \mid \lambda - \alpha_r \in \Lambda(\pi) \}.$$
Clearly, $\Gamma$ is the set of weights of an irreducible representation of $\G(\Delta_r, R)$ corresponding to the same highest weight $\varpi_s$.
The subsystem $\Delta_r$ has type $\rA_{\ell-1}$ in all cases except the last one.
It is also clear that $|\Gamma_0|=1$ for $\Phi=\rA_\ell, \rB_\ell$, $|\Gamma_0|=2$ for $\Phi=\rD_\ell$, $|\Gamma_0|=3$ for $\Phi=\rE_\ell$ and $r=2$.
In the case $\Phi=\rE_\ell$, $r=2$ the subsystem $\Delta_r$ has type $\rD_{\ell-1}$ and $|\Gamma_0|=\ell-1$.

>>>>

\begin{thm}\label{theorem:relative_dv}
The relative elementary subgroup $\E(\Phi, R, I)$ coincides with $A_{rs}$ under the following assumptions on $I$:
\newline \indent \begin{tabular}{l l l l l l}      
\textbullet & $\Phi = \rA_\ell$, $\rB_\ell$, $\ell\geq 2$ & $s = 1$, & $r = \ell$, & $\sr(I) \leq \ell-1$; \\     
\textbullet & $\Phi = \rD_\ell$, $\ell\geq 4$, & $s=1$, & $r=\ell$, & $\sr(I) \leq \ell-2$; \\ 
\textbullet & $\Phi = \rE_\ell$, $\ell=6,7$, & $s=\ell$, & $r=2$, & $\sr(I) \leq \ell-3$; \\ 
\textbullet & $\Phi = \rE_\ell$, $\ell=6,7$, & $s=\ell$, & $r=1$, & $\asr(I)\leq \ell-2$. \\ 
\end{tabular}
\end{thm}
The proof of Theorem~\ref{theorem:relative_dv} occupies the rest of this section.

@@

\begin{proof}[Proof of Theorem~\ref{theorem:relative_dv}]
We first show that under specified assumptions on $I$ one can meet the conditions of Lemma~\ref{lemma:Stein_reduction}.
Consider the following two subsets of $\Lambda(\pi)$:
$$\Gamma = \varpi_s- (\Sigma_s^+\cap \Delta_r),\quad \Gamma_0 = \{\lambda \in \Gamma \mid \lambda - \alpha_r \in \Lambda(\pi) \}.$$
Clearly, $\Gamma$ is the set of weights of an irreducible representation of $\G(\Delta_r, R)$ corresponding to the same highest weight $\varpi_s$.
The subsystem $\Delta_r$ has type $\rA_{\ell-1}$ in all cases except the last one.
It is also clear that $|\Gamma_0|=1$ for $\Phi=\rA_\ell, \rB_\ell$, $|\Gamma_0|=2$ for $\Phi=\rD_\ell$, $|\Gamma_0|=3$ for $\Phi=\rE_\ell$ and $r=2$.
In the case $\Phi=\rE_\ell$, $r=2$ the subsystem $\Delta_r$ has type $\rD_{\ell-1}$ and $|\Gamma_0|=\ell-1$.

### 2015-04-30 paper.tex
Fixes in stability conditions
Rewritten BK and stability conditions

For a representation $\pi$ of $\G(\Phi, R)$ on $V$ we denote by $\Lambda(\pi)$ the set of weights of $\pi$.
It will be convenient to use standard numbering for weights of representation $\pi$ in the case when $\Phi$ is classical and $\pi$ is a vector representation (cp.~\cite[§1B]{St78}):
$$\begin{array}{cll}
  1,2,\ldots, l+1 & \text{in the case} & \Phi =\rA_l, \\
  1,2,\ldots l, 0, -\ell,\ldots, -2, -1 & \text{in the case} & \Phi =\rB_\ell, \\
  1,2,\ldots l, -\ell,\ldots, -2, -1 & \text{in the cases}   & \Phi =\rC_\ell, \rD_\ell. \\
\end{array}$$
In particular, we write $1$ instead of $\varpi_1$, $2$ instead of $\varpi_1-\alpha_1$ and so on.

@@

\begin{rem} Let $\lambda_1, \lambda_2 \in \Lambda(\pi)$ be a pair of weights of a representation $\pi$ of a classical group such that $\lambda_1-\lambda_2\in \Phi$.
In this situation it will be convenient for us to write $t_{\lambda_1,\lambda_2}(\xi)$ instead of $t_{\lambda_1-\lambda_2}(\xi)$.
In particular, for $\Phi=\rA_\ell$ we have $t_{1,2}(\xi)=t_{1-2}(\xi)=t_{\varpi_1 - \varpi_1 + \alpha_1}(\xi) = t_{\alpha_1}(\xi)$. \end{rem}

@@

\section{Relative Bass---Kolster decomposition}\label{sec:bass-kolster}
\input{bass-kolster}

\section{Relative Dennis-Vaserstein decompositions}\label{sec:dennis-vaserstein}
\input{dennis-vaserstein}

>>>>

For a representation $\pi$ of $\G(\Phi, R)$ on $V$ we denote by $\Lambda(\pi)$ the set of weights of $\pi$.
It will be convenient to use standard numbering for weights of representation $\pi$ in the case when $\Phi$ is classical and $\pi$ is a vector representation (cp.~\cite[§1B]{St78}):
$$\begin{array}{cll}
  1,2,\ldots, l+1 & \text{in the case} & \Phi =\rA_l, \\
  1,2,\ldots l, 0, -\ell,\ldots, -2, -1 & \text{in the case} & \Phi =\rB_\ell, \\
  1,2,\ldots l, -\ell,\ldots, -2, -1 & \text{in the cases}   & \Phi =\rC_\ell, \rD_\ell. \\
\end{array}$$
For example, we write $1$ instead of $\varpi_1$, $2$ instead of $\varpi_1-\alpha_1$ and so on.

@@

\begin{rem} Let $\lambda_1, \lambda_2 \in \Lambda(\pi)$ be a pair of weights of a representation $\pi$ of a classical group such that $\lambda_1-\lambda_2\in \Phi$.
In this situation it will be convenient for us to write $t_{\lambda_1,\lambda_2}(\xi)$ instead of $t_{\lambda_1-\lambda_2}(\xi)$.
For example, for $\Phi=\rA_\ell$ we have $t_{1,2}(\xi)=t_{1-2}(\xi)=t_{\varpi_1 - \varpi_1 + \alpha_1}(\xi) = t_{\alpha_1}(\xi)$. \end{rem}

@@

\section{Relative parabolic factorizations}
\subsection{Relative Bass---Kolster decompositions}\label{sec:bass-kolster}
\input{bass-kolster}
\subsection{Relative Dennis-Vaserstein decompositions}\label{sec:dennis-vaserstein}
\input{dennis-vaserstein}

### 2015-04-30 stability-conditions.tex
Rewritten BK and stability conditions

Let $R$ be a commutative ring. We denote by $\Max(R)$ its \emph{maximum spectrum} i.e. the set of maximal ideals of $R$ equipped with Zariski topology.
The following result gives an upper estimate for the absolute stable rank (see~\cite[Theorem~2.3]{EO}, \cite[Theorem~3.7]{MKV}).
\begin{thm}
Let $R$ be a commutative ring whose maximum spectrum $\Max(R)$ can be covered with a finite family of subsets $X_1\ldots, X_m$ such that the topological dimension of every subset $X_i$ does not exceed some number $d$. Then $\asr(A)\leqslant d+1$.
\end{thm}
Clearly one has
\[ \sr(R,I)\leqslant\asr(R,I)\leqslant\asr(R)\leqslant \dim(\Max(R))+1\leqslant\dim(\Spec(R))+1. \]

\subsection{Action of unipotent radicals}
The next lemma is an immediate consequence of the definition of stable rank.
\begin{lemma}\label{lemma:srUnip} Let $v\in\Ums(n, I)$ be a unimodular column. Assume that $\sr(I)\leq n-m$ for some $1\leqslant m < n$.
Then one can find a matrix $a\in \M(n-m,m,I)$ such that the column $(e_{n-m},a)v\in \Ums(n-m, I)$. \end{lemma}

Let $v\in R^{2\ell}$ be a vector of the standard vector representation of $\G(\rD_\ell, R)$.
It will be convenient for us to denote by $v^+$ and $v^-$ the upper and lower halves of $v$ i.e. $v^+=(v_1,\ldots, v_\ell)^t$, $v^-=(v_{-\ell},\ldots, v_{-1})^t$.
\begin{lemma}\label{lemma:asrUnip}
 Let $\Phi=\rD_\ell$ and assume that $\asr(R, I)\leq \ell -1$. Then for any $I$-unimodular column $v=(v^+, v^-)^t\in\Ums(2\ell, R, I)$
 there exist $g\in\U(\Sigma^+_\ell, I)$ such that $(g \cdot v)^+ \in \Ums(\ell, R, I)$.
\end{lemma}
\begin{proof} Denote by $J$ the ideal of $R$ spanned by entries of $v^-$. Clearly, $J \leq I$.
By Proposition~\ref{prop:sr_properties} $\sr(I/J) \leq \ell-1$ and therefore
the elementary group $\E(\rA_{\ell-1}, R/J, I/J)$ acts transitively on $\Ums(\ell, I/J)$ (see~\cite[Theorem~2.3c]{Va69}). 
This implies existence of $h\in \E(\Delta_\ell, R, I)$ such that for $v' = h \cdot v$ one has $(v')^+ \equiv e_1$ modulo $J$.

@@

\begin{lemma} \label{lemma:uraction} 
Let $\Phi=\rA_\ell, \rD_\ell$. Denote by $\pi$ standard vector representation of $\G(\Phi, R)$ on $V=R^n$, $n=\ell+1,2\ell$.
Assume that one of the following conditions holds:
\begin{itemize}
 \item $\Phi=\rA_\ell$, $\Gamma=\{ k+1, \ldots, \ell+1\} \subset \Lambda(\pi)$ and $\sr(I)\leq k\leq \ell$;
 \item $\Phi=\rD_\ell$, $\Gamma=\{-l,\ldots, -2, -1\} \subset \Lambda(\pi)$ and $\asr(R, I)\leq \ell-1$. 
\end{itemize}
Then for any $I$-unimodular column $g\in \G(\Phi, R, I)$ there exist $x\in \U(\Phi^+, I)$, $y\in \U(\Phi^-, I)$ such that $(yxg \cdot v^+)_\lambda = 0$ for $\lambda\in \Gamma$.
\end{lemma}
\begin{proof}
\textsc{Case $\Phi=\rA_\ell$.} \textbf{TODO: Write proof for this case.}

>>>>

Let $R$ be a commutative ring. We denote by $\Max(R)$ its \emph{maximum spectrum} i.e. the set of maximal ideals of $R$ equipped with Zariski topology.
The following result gives an upper estimate for the absolute stable rank (see~\cite[Theorem~2.3]{EO}, \cite[Theorem~3.7]{MKV}).
\begin{thm}
Let $R$ be a commutative ring whose maximum spectrum $\Max(R)$ can be covered with a finite family of subsets 
$X_1\ldots, X_m$ such that the topological dimension of every subset $X_i$ does not exceed some number $d$. 
Then $\asr(A)\leqslant d+1$.
\end{thm}
%TODO: Elaborate
Clearly one has
\[\sr(I)\leqslant\asr(I)\leqslant\asr(R)\leqslant \dim(\Max(R))+1\leqslant\dim(\Spec(R))+1. \]

\subsection{Action of unipotent radicals}
Let $v\in R^{2\ell}$ be a vector of the standard vector representation of $\G(\rD_\ell, R)$.
It will be convenient for us to denote by $v^+$ and $v^-$ the upper and lower halves of $v$ i.e. $v^+=(v_1,\ldots, v_\ell)^t$, $v^-=(v_{-\ell},\ldots, v_{-1})^t$.
\begin{lemma}\label{lemma:asrUnip}
 Let $\Phi=\rD_\ell$ and assume that $\asr(R, I)\leq \ell -1$. Then for any $I$-unimodular column $v=(v^+, v^-)^t\in\Ums(2\ell, R, I)$
 there exist $g\in\U(\Sigma^+_\ell, I)$ such that $(g \cdot v)^+ \in \Ums(\ell, R, I)$.
\end{lemma}
\begin{proof} Denote by $J$ the ideal of $R$ spanned by entries of $v^-$. Clearly, $J \leq I$.
By Proposition~\ref{prop:sr_properties} $\sr(I/J) \leq \ell-1$ and therefore
the elementary group $\E(\rA_{\ell-1}, R/J, I/J)$ acts transitively on $\Ums(\ell, I/J)$ (see~\cite[Theorem~2.3c]{Va69}). 
This implies existence of $h\in \E(\Delta_\ell, R, I)$ such that for $v' = h \cdot v$ one has $(v')^+ \equiv e_1$ modulo $J$.

@@

\begin{lemma} \label{lemma:uraction} 
Let $\Phi=\rA_\ell, \rD_\ell$. Denote by $\pi$ the standard vector representation of $\G(\Phi, R)$ on $V=R^n$, $n=\ell+1,2\ell$.
Assume that one of the following conditions holds:
\begin{itemize}
 \item $\Phi=\rA_\ell$, $\Gamma=\{ k+1, \ldots, \ell+1\} \subset \Lambda(\pi)$ and $\sr(I)\leq k\leq \ell$;
 \item $\Phi=\rD_\ell$, $\Gamma=\{-l,\ldots, -2, -1\} \subset \Lambda(\pi)$ and $\asr(I)\leq \ell-1$. 
\end{itemize}
Then for $g\in \G(\Phi, R, I)$ there exist $x\in \U(\Phi^+, I)$, $y\in \U(\Phi^-, I)$ such that $(yxg \cdot v^+)_\lambda = 0$ for $\lambda\in \Gamma$.
\end{lemma}
\begin{proof} Denote by $v$ the image of the highest weight vector $v^+$ under $g$.

\textsc{Case $\Phi=\rA_\ell$.} From the definition of the relative stable rank it follows that we can find 
$x= \left(\begin{smallmatrix} e_k & a \\ * & e_{n-k} \end{smallmatrix}\right) \in \U(\Sigma_k^+, I)$ such that 
the upper $k$ components of $v'= x \cdot v$ form an $I$-unimodular column. 
Now, to obtain zeroes at desired positions it remains to subtract from $v'_{k+1},\ldots, v'_{\ell+1}$ suitable multiples of $v'_1,\ldots v'_k$.
Clearly, this operation corresponds to the left multiplication by some element $y\in\U(\Sigma_k^-, I)$.
